---
title: "ISSS626 Hands On Exercise 7"
---

# **Calibrating Hedonic Pricing Model for Private Highrise Property with GWR Method**

## Overview

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, you will learn how to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp)models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

```{r}
setwd('/Users/zishengwong/R Coding/ISSS626 Geospatial Analytics/Hands-on_Ex/Hands-on_ex07')
```

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, 
               sf, sfdep, GWmodel, tmap, 
               tidyverse, gtsummary,
               performance, RColorBrewer)
```

## Importing data

```{r}
mpsz = st_read(dsn = "./data/geospatial", 
               layer = "MP14_SUBZONE_WEB_PL")
```

## Geospatial Data Wrangling

### Updating CRS information

```{r}
mpsz_svy21 <- st_transform(mpsz, 3414)
```

```{r}
st_crs(mpsz_svy21)
```

```{r}
st_bbox(mpsz_svy21)
```

## Aspatial Data Wrangling

### Importing data

```{r}
condo_resale = read_csv(
  "./data/aspatial/Condo_resale_2015.csv")
```

```{r}
glimpse(condo_resale)
```

```{r}
head(condo_resale$LONGITUDE)
```

```{r}
head(condo_resale$LATITUDE)
```

```{r}
summary(condo_resale)
```

### Converting aspatial data frame into a sf object

```{r}
condo_resale.sf <- st_as_sf(condo_resale,
                            coords = c("LONGITUDE", "LATITUDE"),
                            crs=4326) %>%
  st_transform(crs=3414)
```

```{r}
head(condo_resale.sf)
```

## Exploratory Data Analysis

### EDA using statistical graphics

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

The figure above reveals a right skewed distribution. This means that more condominium units were transacted at relative lower prices.

Statistically, the skewed dsitribution can be normalised by using log transformation. The code chunk below is used to derive a new variable called *LOG_SELLING_PRICE* by using a log transformation on the variable *SELLING_PRICE*. It is performed using `mutate()` of **dplyr** package.

```{r}
condo_resale.sf <- condo_resale.sf %>%
  mutate(`LOG_SELLING_PRICE` = log(SELLING_PRICE))
```

```{r}
ggplot(data=condo_resale.sf, 
       aes(x=`LOG_SELLING_PRICE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")
```

### Multiple histogram plots

```{r}
AREA_SQM <- ggplot(data=condo_resale.sf, 
                   aes(x= `AREA_SQM`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

AGE <- ggplot(data=condo_resale.sf, 
              aes(x= `AGE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_CBD <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_CBD`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_CHILDCARE <- ggplot(data=condo_resale.sf, 
                         aes(x= `PROX_CHILDCARE`)) + 
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_ELDERLYCARE <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_ELDERLYCARE`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_URA_GROWTH_AREA <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_URA_GROWTH_AREA`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_HAWKER_MARKET <- ggplot(data=condo_resale.sf, 
                             aes(x= `PROX_HAWKER_MARKET`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_KINDERGARTEN <- ggplot(data=condo_resale.sf, 
                            aes(x= `PROX_KINDERGARTEN`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_MRT <- ggplot(data=condo_resale.sf, 
                   aes(x= `PROX_MRT`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_PARK <- ggplot(data=condo_resale.sf, 
                    aes(x= `PROX_PARK`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                           aes(x= `PROX_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

PROX_TOP_PRIMARY_SCH <- ggplot(data=condo_resale.sf, 
                               aes(x= `PROX_TOP_PRIMARY_SCH`)) +
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue")

ggarrange(AREA_SQM, AGE, PROX_CBD, PROX_CHILDCARE, 
          PROX_ELDERLYCARE, PROX_URA_GROWTH_AREA,
          PROX_HAWKER_MARKET, PROX_KINDERGARTEN, PROX_MRT,
          PROX_PARK, PROX_PRIMARY_SCH, PROX_TOP_PRIMARY_SCH,  
          ncol = 3, nrow = 4)
```

### Drawing Statitical Point Map

```{r}
tmap_mode("view")
```

```{r}
tm_shape(mpsz_svy21)+
  tm_polygons() +
tm_shape(condo_resale.sf) +  
  tm_dots(fill = "SELLING_PRICE",
          fill_alpha = 0.6,
          size = 0.5,
          fill.scale = tm_scale_intervals(
            style="quantile")) +
  tm_view(set_zoom_limits = c(11,14))
```

```{r}
tmap_mode("plot")
```

## Hedonic Pricing Modelling in R

### Simple Linear Regression Method

```{r}
condo.slr <- lm(formula=SELLING_PRICE ~ AREA_SQM, 
                data = condo_resale.sf)
```

```{r}
summary(condo.slr)
```

The output report reveals that the SELLING_PRICE can be explained by using the formula:

```         
      *y = -258121.1 + 14719x1*
```

The R-squared of 0.4518 reveals that the simple regression model built is able to explain about 45% of the resale prices.

Since p-value is much smaller than 0.0001, we will reject the null hypothesis that mean is a good estimator of SELLING_PRICE. This will allow us to infer that simple linear regression model above is a good estimator of *SELLING_PRICE*.

The **Coefficients:** section of the report reveals that the p-values of both the estimates of the Intercept and ARA_SQM are smaller than 0.001. In view of this, the null hypothesis of the B0 and B1 are equal to 0 will be rejected. As a results, we will be able to infer that the B0 and B1 are good parameter estimates.

To visualise the best fit curve on a scatterplot, we can incorporate `lm()` as a method function in ggplot’s geometry as shown in the code chunk below.

```{r}
ggplot(data=condo_resale.sf,  
       aes(x=`AREA_SQM`, y=`SELLING_PRICE`)) +
  geom_point() +
  geom_smooth(method = lm)
```

### Multiple Linear Regression Method

Before building a multiple regression model, it is important to ensure that the indepdent variables used are not highly correlated to each other. If these highly correlated independent variables are used in building a regression model by mistake, the quality of the model will be compromised. This phenomenon is known as **multicollinearity** in statistics.

Correlation matrix is commonly used to visualise the relationships between the independent variables. Beside the `pairs()` of R, there are many packages support the display of a correlation matrix. In this section, the [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) package will be used.

The code chunk below is used to plot a scatterplot matrix of the relationship between the independent variables in *condo_resale* data.frame.

```{r}
corrplot(cor(condo_resale[, 5:23]), 
         diag = FALSE, 
         order = "AOE",
         tl.pos = "td", 
         tl.cex = 0.5, 
         method = "number", 
         type = "upper")
```

Matrix reorder is very important for mining the hiden structure and patter in the matrix. There are four methods in corrplot (parameter order), named “AOE”, “FPC”, “hclust”, “alphabet”. In the code chunk above, AOE order is used. It orders the variables by using the *angular order of the eigenvectors* method suggested by [Michael Friendly](https://www.datavis.ca/papers/corrgram.pdf).

From the scatterplot matrix, it is clear that ***Freehold*** is highly correlated to ***LEASE_99YEAR***. In view of this, it is wiser to only include either one of them in the subsequent model building. As a result, ***LEASE_99YEAR*** is excluded in the subsequent model building.

#### Building a hedonic pricing model using multiple linear regression

```{r}
condo.mlr <- lm(
  formula = SELLING_PRICE ~ AREA_SQM + AGE  + 
    PROX_CBD + PROX_CHILDCARE + PROX_ELDERLYCARE +
    PROX_URA_GROWTH_AREA + PROX_HAWKER_MARKET   +
    PROX_KINDERGARTEN   + PROX_MRT  + PROX_PARK +
    PROX_PRIMARY_SCH + PROX_TOP_PRIMARY_SCH +
    PROX_SHOPPING_MALL  + PROX_SUPERMARKET + 
    PROX_BUS_STOP   + NO_Of_UNITS + 
    FAMILY_FRIENDLY + FREEHOLD, 
  data=condo_resale.sf)
summary(condo.mlr)
```
