---
title: "Take_Home_Ex03"
---

# Setting the Scene

Housing is an essential component of household wealth worldwide. Purchasing a home has always represented a major investment for most individuals. The price of housing is influenced by a wide range of factors. Some are macro-level, such as the overall state of the economy or the inflation rate, while others are property-specific. These can be further categorized into structural and locational factors.

Structural factors refer to the characteristics of the property itself, such as size, fittings, and tenure. Locational factors, on the other hand, relate to the neighbourhood context of the property, including proximity to childcare centres, public transport, and shopping facilities.

Traditionally, housing resale price models have been estimated using the Ordinary Least Squares (OLS) method. However, this approach does not account for spatial autocorrelation and spatial heterogeneity, which are commonly present in geographic datasets such as housing transactions. When spatial autocorrelation exists, OLS estimation of housing resale price models may produce biased, inconsistent, or inefficient results (Anselin, 1998). To address these limitations, **Geographically Weighted Models (GWMs)** have been introduced to more accurately calibrate both explanatory and predictive models for housing resale prices.

# Objective

The objective of this take-home exercise is to develop a spatially informed model to determine factors affecting or to predict HDB resale prices in Singapore, making use of appropriate geospatial analytics methods.

```{r}
setwd('/Users/zishengwong/R Coding/ISSS626 Geospatial Analytics/Take Home Exercise/Take Home Exercise 3')
```

```{r}
pacman::p_load(httr, tidyverse, sf, tmap, 
               jsonlite, progress)

```

```{r}
pacman::p_load(sf, spdep, GWmodel, SpatialML, 
               tmap, rsample, Metrics, tidyverse)
```

# Geocoding HDB resale data

```{r}
HDBresale <- read_csv(
  "data/rawdata/HDBresale.csv") %>%
  filter(month >= "2024-07" & month <= "2025-06")
```

-   Combining the block and street name into an address. The addresses will be used to geocode instead of postal code that you learned in In-class Exercise 3b,

-   Passing the address as the **searchVal** in our query,

-   Sending the query to **OneMapSG search**. Note: Since we don’t need all the address details, we can set getAddrDetails as ‘N’,

-   Converting response (JSON object) to text,

-   Saving response in text form as a dataframe, and

-   Retain the latitude and longitude for our output.

```{r}
geocode <- function(block, streetname) {
  base_url <- "https://onemap.gov.sg/api/common/elastic/search"
  address <- paste(block, streetname, sep = " ")
  query <- list("searchVal" = address, 
                "returnGeom" = "Y",
                "getAddrDetails" = "N",
                "pageNum" = "1")
  
  res <- GET(base_url, query = query)
  restext<-content(res, as="text")
  
  output <- fromJSON(restext)  %>% 
    as.data.frame %>%
    select(results.LATITUDE, results.LONGITUDE)

  return(output)
}
```

```{r}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false

HDBresale$LATITUDE <- 0
HDBresale$LONGITUDE <- 0

for (i in 1:nrow(HDBresale)) {
  temp_output <- geocode(HDBresale[i, 4], HDBresale[i, 5])
  HDBresale$LATITUDE[i] <- temp_output$results.LATITUDE
  HDBresale$LONGITUDE[i] <- temp_output$results.LONGITUDE
}
```

## Proximity Analysis

Convert HDBresale to sf object

```{r}
HDBresale_sf <- HDBresale %>%
  st_as_sf(
    coords = c("LONGITUDE", "LATITUDE"),
    crs=4326) %>%
  st_transform(crs = 3414)
```

### Preschool

```{r}
preschool = st_read(
  "data/rawdata/PreSchoolsLocation.kml") %>%
  st_transform(crs= 3414)
```

```{r}
within_350m_preschool <- st_is_within_distance(
  HDBresale_sf, preschool, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_PRESCHOOL = map_int(within_350m_preschool, length))
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_PRESCHOOL = as.numeric(
      st_distance(geometry,
                  preschool[st_nearest_feature(
                    geometry, preschool), ], 
                  by_element = TRUE)
      )
    )
```

### Eldercare

```{r}
eldercare = st_read(
  "./data/EldercareServicesKML.kml") %>%
  st_transform(crs= 3414)
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_ELDERCARE = as.numeric(
      st_distance(geometry,
                  eldercare[st_nearest_feature(
                    geometry, eldercare), ], 
                  by_element = TRUE)
      )
    )
```

### Hawker center

```{r}
hawkercenter = st_read(
  "./data/HawkerCentresKML.kml") %>%
  st_transform(crs= 3414)
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_HAWKER = as.numeric(
      st_distance(geometry,
                  hawkercenter[st_nearest_feature(
                    geometry, hawkercenter), ], 
                  by_element = TRUE)
      )
    )
```

### SuperMarkets

```{r}
supermarkets = st_read(
  "./data/SupermarketsKML.kml") %>%
  st_transform(crs= 3414)
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_SUPERMARKETS = as.numeric(
      st_distance(geometry,
                  supermarkets[st_nearest_feature(
                    geometry, supermarkets), ], 
                  by_element = TRUE)
      )
    )
```

### Busstops

```{r}
BusStop = st_read(dsn = "data/LTADataMall/", 
                  layer = "BusStop") %>%
  st_transform(crs = 3414)
```

```{r}
within_350m_bus <- st_is_within_distance(
  HDBresale_sf, BusStop, dist = 350)

HDBresale_sf <- HDBresale_sf %>%
  mutate(WITHIN_350M_BUSSTOP = map_int(within_350m_bus, length))
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_BUSSTOP = as.numeric(
      st_distance(geometry,
                  BusStop[st_nearest_feature(
                    geometry, BusStop), ], 
                  by_element = TRUE)
      )
    )
```

### Train Station

```{r}
Sys.setenv(OGR_GEOMETRY_ACCEPT_UNCLOSED_RING = "NO")

TrainStation <- st_read(
  dsn = "data/TrainStation/",
  layer = "RapidTransitSystemStation"
) %>%
  st_transform(crs = 3414)
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_TRAINSTATION = as.numeric(
      st_distance(geometry,
                  TrainStation[st_nearest_feature(
                    geometry, TrainStation), ], 
                  by_element = TRUE)
      )
    )
```

### Primary School

```{r}
PrimarySchool <- read_csv("data/Generalinformationofschools.csv")
```

```{r}
geocode_postal <- function(postal_code) {
  base_url <- "https://www.onemap.gov.sg/api/common/elastic/search"
  query <- list(
    searchVal = postal_code,
    returnGeom = "Y",
    getAddrDetails = "N",
    pageNum = "1"
  )
  
  res <- try(httr::GET(base_url, query = query), silent = TRUE)
  if (inherits(res, "try-error")) return(c(NA, NA))
  
  restext <- httr::content(res, as = "text", encoding = "UTF-8")
  
  # parse JSON safely
  output <- try(jsonlite::fromJSON(restext), silent = TRUE)
  if (inherits(output, "try-error")) return(c(NA, NA))
  
  # Check if results exist and are a list
  if (!is.null(output$results) && length(output$results) > 0) {
    first_result <- output$results[[1]]
    if (is.list(first_result) && all(c("LATITUDE", "LONGITUDE") %in% names(first_result))) {
      lat <- as.numeric(first_result$LATITUDE)
      lon <- as.numeric(first_result$LONGITUDE)
      return(c(lat, lon))
    }
  }
  
  # fallback if results missing or malformed
  return(c(NA, NA))
}
```

```{r}
library(dplyr)
library(purrr)
library(tibble)

coords <- map_dfr(PrimarySchool$postal_code, function(code) {
  Sys.sleep(0.5)  # polite delay to avoid rate limits
  latlon <- geocode_postal(code)
  tibble(LATITUDE = latlon[1], LONGITUDE = latlon[2])
})

PrimarySchool <- bind_cols(PrimarySchool, coords)
```

```{r}
PrimarySchool_sf <- PrimarySchool %>%
  filter(!is.na(LATITUDE) & !is.na(LONGITUDE)) %>%
  st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = 4326) %>%
  st_transform(crs = 3414)
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    PROX_PRIMARYSCHOOL = as.numeric(
      st_distance(
        geometry,
        PrimarySchool_sf[st_nearest_feature(geometry, PrimarySchool_sf), ],
        by_element = TRUE
      )
    )
  )
```

Preprocess columns

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(
    STOREY_MID = str_extract(storey_range, "^\\d+") %>% as.numeric() +
                 (str_extract(storey_range, "\\d+$") %>% as.numeric() -
                  str_extract(storey_range, "^\\d+") %>% as.numeric()) / 2
  )

HDBresale_sf <- HDBresale_sf %>%
  mutate(
    # Extract number of years
    lease_years = str_extract(remaining_lease, "\\d+(?= years)") %>% as.numeric() %>% replace_na(0),
    # Extract number of months
    lease_months = str_extract(remaining_lease, "\\d+(?= months)") %>% as.numeric() %>% replace_na(0),
    # Compute total months
    remaining_lease_mths = lease_years * 12 + lease_months
  ) %>%
  select(-lease_years, -lease_months)  # optional: drop intermediate columns
```

```{r}
HDBresale_sf <- HDBresale_sf %>%
  mutate(across(c(PROX_PRESCHOOL,PROX_ELDERCARE, PROX_HAWKER, PROX_TRAINSTATION,
                  PROX_SUPERMARKETS,PROX_BUSSTOP,PROX_PRIMARYSCHOOL),
                ~ . / 1000))  # convert meters to km
```

```{r}
summary(HDBresale_sf)
```

Primary school has entirely NaNs, we will remove it from analysis.

# Correlation analysis

```{r}
HDBresale_sf_copy <- HDBresale_sf  # simple deep copy
```

```{r}
numeric_cols <- HDBresale_sf_copy %>% st_drop_geometry() %>%
  select(where(is.numeric)) %>%
  select(where(~ !all(is.na(.)) & var(., na.rm = TRUE) > 0))
```

```{r}
library(corrplot)
corrplot(
  cor(numeric_cols, use = "pairwise.complete.obs"),  # handle NA
  diag = FALSE, 
  order = "AOE",
  tl.pos = "td", 
  tl.cex = 0.5, 
  method = "number", 
  type = "upper"
)
```

remaining_lease_mths and lease_commence_data have high collinearity. However we will only be using remaining_lease_mths hence it is not a problem.

No other columns have high collinearlity, we will include all in analysis.

# GWR method

## Computing adaptive bandwidth

```{r}
bw_adaptive <- bw.gwr(
  formula = resale_price ~ floor_area_sqm +
    STOREY_MID + remaining_lease_mths + PROX_ELDERCARE + PROX_HAWKER + PROX_PRESCHOOL + PROX_BUSSTOP+ 
    PROX_TRAINSTATION + PROX_SUPERMARKETS + WITHIN_350M_BUSSTOP + WITHIN_350M_PRESCHOOL,
  data=HDBresale_sf,
  approach="CV",
  kernel="gaussian",
  adaptive=TRUE,
  longlat=FALSE)
```

```{r}
write_rds(bw_adaptive, "data/model/bw_adaptive.rds")
```

```{r}
bw_adaptive <- read_rds("data/model/bw_adaptive.rds")
```

## Constructing adaptive bandwidth gwr model

```{r}
gwr_adaptive <- gwr.basic(
  formula = resale_price ~ floor_area_sqm +
    STOREY_MID + remaining_lease_mths + PROX_ELDERCARE + PROX_HAWKER + PROX_PRESCHOOL + PROX_BUSSTOP+ 
    PROX_TRAINSTATION + PROX_SUPERMARKETS + WITHIN_350M_BUSSTOP + WITHIN_350M_PRESCHOOL,
  data=HDBresale_sf,
  bw=bw_adaptive, 
  kernel = 'gaussian', 
  adaptive=TRUE,
  longlat = FALSE)
```

```{r}
write_rds(gwr_adaptive, "data/model/gwr_adaptive.rds")
```

```{r}
gwr_adaptive <- read_rds("data/model/gwr_adaptive.rds")
gwr_adaptive
```

Results above show R square value of 0.723, AIC of 703701 and F-statistic that indicates highly significant (p\<0.001). Overall the OLS model provides a strong baseline, explaining approximately 72% of price variation.
